---
title: "Extract Sargassum Data from OceanCleaner KMLs"
author: "Allison Bailey"
date: "9/17/2019"
output: html_document
---

This notebook is to extract Sargassum data downloaded from Ocean Cleaner's website
2018: http://sargassummonitoring.com/2018-sargassum-sargasses-sargazo/
current (2019): http://sargassummonitoring.com/

Data were downloaded as KML files from the website

Data are stored as points.  
One file per year for presence and one for absence observations

Convert to GeoPackage layers

# Setup
### Load Packages..

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(sf)
library(tidyverse)
library(lubridate)
```

Read KML Data and extract some info into new columns
```{r kml_data}
# https://ryanpeek.github.io/mapping-in-R-workshop/spatial_kml.html
# Read data from KML
presence_2018_kml <- st_read(here::here("data/oceancleaner","2018 Beaches with sargassum seaweeds - Plages avec algues sargasses.kml"))
absence_2018_kml <- st_read(here::here("data/oceancleaner","2018 - Beaches without sargassum - Plages sans sargasses - Playas sin sargazo.kml"))

# Pattern for dates in the format of DD/MM/YYYY or MM/DD/YYYY or MM/YYYY
#  Does not check to confirm that they are valid dates, just the right format

date_pattern <- "([0-9]{2}/)?[0-9]{2}/[0-9]{4}"  # Any of the formats
date_dmy <- "[0-3][0-9]/[0-1][0-9]/[0-9]{4}"  # DD/MM/YYYY (or MM/DD/YY for some dates)
date_my <- "[0-1][0-9]/[0-9]{4}"  # MM/YYYY

# Creates a new column each time, so 2nd statement overwrites the first
# presence_2018_kml$source_date <- dmy(str_extract(presence_2018_kml$Name, date_pattern))
# presence_2018_kml$source_date <- myd(str_extract(presence_2018_kml$Name, date_pattern), truncated = 1)

# Calculate source_date column based on type of date format found
presence_2018 <- presence_2018_kml %>%
  transmute(
    # rename KML default column names
    name_kml = Name,
    desc_kml = Description
  ) %>%
  mutate(
    # Extract any date values from the Name column
    source_date = case_when(
      str_detect(name_kml, date_dmy) ~ dmy(str_extract(name_kml, date_dmy)),
      str_detect(name_kml, date_my) ~ myd(str_extract(name_kml, date_my), truncated = 1)
    )
  ) %>%
  mutate(
    # Create new column to capture the date type
    date_type = case_when(
      str_detect(name_kml, date_dmy) ~ "DMY",
      str_detect(name_kml, date_my) ~ "MY"   
    )
  ) %>%
  mutate(
    # Column for sargassum occurrence (present/absent)
    occur = "present"
  ) %>%
  filter(
    str_detect(name_kml, "FOR 2019", negate = TRUE)
  )

```
Quick Map to Review data
```{r map_it}
# simple map to visualize data
ggplot() +
  labs(x="Longitude (WGS84)", y="Latitude",
       title="Sargassum Locations 2018") + 
  geom_sf(data=presence_2018, col="blue", lwd=0.4, pch=21) +
  theme_bw()
```

Output to Geopackage
```{r geopkg_it}
gpkg_file = here::here("data/oceancleaner","sargassum_oceancleaner.gpkg")
st_write(presence_2018, dsn=gpkg_file, layer='sargassum_present2018', layer_options = "OVERWRITE = YES")

# Try reading it back
# check available layers from a geopackage
st_layers(gpkg_file)

```